Parsing Postfix log files

### 2009-03-03 doc/logparser-architecture.tex
Wrote the first half of the rule conditions section; it needs to be augmented with an example, and maybe some more explanation.

Actions may return a modified input line to the framework that will be
parsed as if read from the input stream, allowing for a simplified version
of cascaded parsing~\cite{cascaded-parsing}.  This powerful facility allows
several rules and actions to parse a single input, potentially simplifying
both rules and actions.  XXX EXPAND WITH CHEMICAL FORMULAE EXAMPLE\@.

@@

A more complicated example is determining the weight of a compound given
its chemical formula.  A chemical formula is a sequence of one or two
character chemical element abbreviations, each optionally followed by a
number indicating how many of that element are present in the formula (if
omitted the number is assumed to be one).  Some parts of a formula may be
repeated: they are enclosed in parentheses, and the closing parenthesis is
followed by a number indicating how many are present in the formula.  A
monomer is a simple compound, i.e.\ it does not have any repeated portions;
compounds with repeated portions are called polymers.  E.g.\ glucose (a
monomer) has the chemical formula $C_{6}H_{12}O_{6}$ and hydrocarbon (a
polymer) is $CH_{3}{(CH_{2})}_{50}CH_{3}$.  The atomic weight of an
individual element is given in periodic tables, but one element in
particular will be important in the later explanation: Hydrogen,
abbreviated H, has an atomic weight of 1.  The weight of a given compound
is the sum of the atomic weights of each of the elements multiplied by the
number of times they occur; this is simple to calculate for a monomer, but
it is harder for a polymer because the weight of the repeated monomer must
be multiplied by the number of times the monomer is present in the compound
(e.g.\ for hydrocarbon the weight of $CH_{2}$ must be multiplied by $50$).
Calculating the weight of a monomer requires a single rule and action: the
action separates each element in the monomer, looks up its weight,
multiplies it by the number of times it is present (one if unspecified),
and adds up the results.  The rule's regex is \verb!^([A-Z]([a-z])?\d*)+$!,
which matches:

@@

The regex does not guarantee that the chemical element abbreviations in the
formula are correct; the regex could be modified to check that by replacing
\verb![A-Z]([a-z])?! with all the valid abbreviations, simplifying the
action at the cost of complicating the regex (which could be automatically
generated).

Parsing a polymer is more difficult because of the repeated monomers: each
monomer's weight needs to be calculated, multiplied by the number of times
it is present in the compound, and that figure added to the sum.  XXX CAN
POLYMERS BE NESTED\@?  Without cascaded parsing the action would need to
maintain state when recursively parsing the monomers embedded in the
polymer.  With cascaded parsing available three rules and actions could be
used as shown in \tableref{Regexes and Actions to calculate the weight of a
chemical compound}:

\newlength{\regexwidth}
% This figure from experimentation; set the length to 0pt, see how big a
% hbox latex complains about, and then set the length.  Stupid latex cannot
% do \settowidth{\verb}.  Poo.
\setlength{\regexwidth}{130pt}
\newlength{\actionwidth}
\setlength{\actionwidth}{\textwidth}
\addtolength{\actionwidth}{-\regexwidth}
% This is the width of the number column, roughly.
\addtolength{\actionwidth}{-30pt}
% This figure also from experimentation.
\addtolength{\actionwidth}{-24pt}
\begin{table}[ht]
    \caption{Regexes and Actions to calculate the weight of a chemical
    compound}
    \empty{}\label{Regexes and Actions to calculate the weight of a
    chemical compound}
    \begin{tabular}{lp{\regexwidth}p{\actionwidth}}

        \tabletopline{}%
        No.   & Regex                       & Action \\
        \tablemiddleline{}%
        1     & \verb'((?!H\d+([()]|$))'    & Parse a monomer as described above,  \\
              & \verb' ([A-Z]([a-z])?\d*)+)'& returning \verb!Hnnn!, where \verb!nnn! is
                                              the weight of the monomer.  The first part
                                              of the regex prevents it from parsing what
                                              it returns (\verb!Hnnn!) and the parsing
                                              process getting stuck in a loop.  \\
        2     & \verb!\(H(\d+)\)(\d+)!      & Parse a repeated monomer that has been
                                              rewritten by the first rule and action,
                                              returning \verb!Hnnn!, where \verb!nnn! is
                                              the product of the first and second numbers
                                              captured by the regex.  \\
        3     & \verb!^H(\d+)+$!            & Repeated applications of rules and
                                              actions one and two will have rewritten the
                                              input to a sequence of one or more
                                              \verb!Hnnn!, where \verb!nnn! is the weight
                                              of the compound.  This action should do
                                              whatever is required with the calculated
                                              weight.  \\
        \tablebottomline{}%

@@

XXX ARE THERE ANY PROBLEMS TO BE ADDRESSED HERE\@?  I DO NOT THINK THAT I
NEED TO WORRY ABOUT ORDERING OR PRIORITY, BUT I SHOULD REVIEW TO BE SURE\@.

@@

Decoupling the rules from the actions and framework enables other rule
management approaches to be used, e.g.\ instead of manually editing
existing rules or adding new rules, machine learning techniques could be
used to automate the process.  If this approach was taken the choice of
machine learning technique would be constrained by the size of typical data
sets (see \sectionref{parser efficiency}).  Techniques requiring the full
data set when training would be impractical; Instance Based
Learning~\cite{instance-based-learning} techniques that automatically
determine which inputs from the training set are valuable and which inputs
can be discarded might reduce the data required to a manageable size.  A
parser could also dynamically create new rules in response to certain
inputs, e.g.\ XXX GIVE A GOOD EXAMPLE\@.  These avenues of research and
development have not been pursued by the author, but the architecture
allows them to easily be undertaken independently.

@@

XXX ATTACHING CONDITIONS TO RULES\@; GIVE PLP EXAMPLE AND REFER TO
IMPLEMENTATION AND RESULTS SECTIONS\@?

>>>>

Actions may return a modified input line to the framework that will be
parsed as if read from the input stream, allowing for a simplified version
of cascaded parsing~\cite{cascaded-parsing}.  This powerful facility allows
several rules and actions to parse a single input, potentially simplifying
both rules and actions.

@@

A more complicated example is determining the weight of a compound given
its chemical formula.  A chemical formula is a sequence of one or two
character chemical element abbreviations, each optionally followed by a
number indicating how many of that element are present in the formula (if
omitted the number is assumed to be one).  Some parts of a formula may be
repeated: they are enclosed in parentheses, and the closing parenthesis is
followed by a number indicating how many are present in the formula.  A
monomer is a simple compound, i.e.\ it does not have any repeated portions;
compounds with repeated portions are called polymers.  E.g.\ glucose (a
monomer) has the chemical formula $C_{6}H_{12}O_{6}$ and hydrocarbon (a
polymer) is $CH_{3}{(CH_{2})}_{50}CH_{3}$.  The atomic weight of an
individual element is given in periodic tables, but one element in
particular will be important in the later explanation: Hydrogen,
abbreviated H, has an atomic weight of 1.  The weight of a given compound
is the sum of the atomic weights of each of the elements multiplied by the
number of times they occur; this is simple to calculate for a monomer, but
it is harder for a polymer because the weight of the repeated monomer must
be multiplied by the number of times the monomer is present in the compound
(e.g.\ for hydrocarbon the weight of $CH_{2}$ must be multiplied by $50$).

Calculating the weight of a monomer requires a single rule and action: the
action separates each element in the monomer, looks up its weight,
multiplies it by the number of times it is present (one if unspecified),
and adds up the results.  The rule's regex is \verb!^([A-Z]([a-z])?\d*)+$!,
which matches:

@@

The regex does not guarantee that the chemical element abbreviations in the
formula are correct; the regex could be modified to check that by replacing
\verb![A-Z]([a-z])?! with all the valid abbreviations, slightly simplifying
the action at the cost of complicating the regex (which could be
automatically generated).

Parsing a polymer is more difficult because of the repeated monomers: each
monomer's weight needs to be calculated, multiplied by the number of times
it is present in the compound, and that figure added to the sum.  XXX CAN
POLYMERS BE NESTED\@?  Without cascaded parsing the action would need to
maintain state while recursively parsing the monomers embedded in the
polymer.  With cascaded parsing available three rules and actions could be
used as shown in \tableref{Regexes and Actions to calculate the weight of a
chemical compound}:

\newlength{\regexwidth}
% These figures come from experimentation; set the length to 0pt, see how
% big a hbox latex complains about, and then set the length.  Stupid latex
% cannot do \settowidth{\verb}.  Poo.
\setlength{\regexwidth}{130pt}
\newlength{\actionwidth}
\setlength{\actionwidth}{\textwidth}
\addtolength{\actionwidth}{-\regexwidth}
% This is the width of the number column, roughly.
\addtolength{\actionwidth}{-30pt}
% This is the overhead of space around columns, on either side, etc.
\addtolength{\actionwidth}{-24pt}

\begin{table}[ht]
    \caption{Regexes and Actions to calculate the weight of a chemical
    compound}
    \empty{}\label{Regexes and Actions to calculate the weight of a
    chemical compound}
    \begin{tabular}{lp{\regexwidth}p{\actionwidth}}

        \tabletopline{}%
        No.   & Regex                       & Action \\
        \tablemiddleline{}%
        1     & \verb'((?!H\d+([()]|$))'    & Parse a monomer as described above,  \\
              & \verb' ([A-Z]([a-z])?\d*)+)'& returning \verb!Hnnn!, where \verb!nnn! is
                                              the weight of the monomer.  The first part
                                              of the regex prevents it from parsing the
                                              return value of this action (\verb!Hnnn!),
                                              so the parsing process does not get stuck in
                                              a loop.  \\
        2     & \verb!\(H(\d+)\)(\d+)!      & Parse a repeated monomer that has been
                                              rewritten by the first rule and action,
                                              returning \verb!Hnnn!, where \verb!nnn! is
                                              the product of the first and second numbers
                                              captured by the regex.  \\
        3     & \verb!^H(\d+)+$!            & Repeated applications of rules and
                                              actions one and two will have rewritten the
                                              input to \verb!Hnnn!, where \verb!nnn! is
                                              the weight of the compound.  This action
                                              should do whatever is required with the
                                              calculated weight.  \\
        \tablebottomline{}%

@@

@@

Decoupling the rules from the actions and framework enables other rule
management approaches to be used, e.g.\ instead of manually editing
existing rules or adding new rules, machine learning techniques could be
used to automate the process.  If this approach was taken the choice of
machine learning technique would be constrained by the size of typical data
sets (see \sectionref{parser efficiency}).  Techniques requiring the full
data set when training would be impractical; Instance Based
Learning~\cite{instance-based-learning} techniques that automatically
determine which inputs from the training set are valuable and which inputs
can be discarded might reduce the data required to a manageable size.  A
parser could also dynamically create new rules in response to certain
inputs, e.g.\ parsing a subroutine declaration could cause a rule to be
created that parses calls to that subroutine, checking that the arguments
used agree with the subroutine's signature.

These avenues of research and development have not been pursued by the
author, but the architecture allows them to easily be undertaken
independently.

@@

Rules can specify conditions that will be evaluated by the framework before
attempting to use a rule to recognise an input: if the condition evaluates
to true the rule will be used, if not the rule will be skipped.  Conditions
can be as simple or complex as the parser requires, though naturally as the
complexity rises so does the difficulty in understanding how everything
interacts.  The framework has to evaluate each condition, so as the
complexity of conditions increases so does the complexity of the code
required to evaluate them; beyond a certain level of complexity conditions
should probably be written in a proper programming language, e.g.\ taking
advantage of dynamic languages' support for evaluating code at run-time, or
embedding a language like Lua
(\urlLastCheckedNoParens{http://ww.lua.org/}{2009/03/03}).  If an
implementation is going to use conditions so complex that they will require
a full programming language to implement them, the design may need to be
re-thought, including the choice to use this architecture --- there may be
other architectures more suitable.

Conditions that examine the input will be the easiest to understand,
because they can be understood in isolation --- they do not depend on
variables set by actions or other rules.  Conditions that examine the input
may be complex, but simple conditions can be quite useful too, e.g.\ every
Postfix log \empty{}line contains the name of the Postfix component that
produced it, so every rule used in \parsername{} specifies the component
whose log \empty{}lines it recognises as a condition, reducing the number
of rules that will be used when classifying an input (see \sectionref{rules
in implementation} for details).

Conditions can also check the value of variables that have been set by
either actions or rules; it will probably be easier to understand how a
variable's value will be used and changed if it is set by rules only rather
than by actions, because the chain of checking and setting can be followed
from rule to rule.  The downside to actions setting variables used in rule
conditions is action at a distance: to understand when a rule's condition
will be true requires understanding not just every rule but also every
action.  The framework probably does not need to support the same level of
complexity and flexibility when setting variables from rules (actions can
obviously use complex code if they are setting variables); however, if the
framework allows complex conditions, that code can probably be easily
extended to support complex variable assignments too.
